ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-08-14 paper/EPB-Final-Submission.tex
adjusting tables for publication

\corrauth{Bruno Dias dos Santos}

@@

The objective of this paper is to introduce \{ActiveCA\}, an open data
product with data from all Time Use Survey (TUS) cycles of the General
Social Surveys (GSSs). Open data products (ODPs) are the outcome of a
process that transforms raw data (open or not) into analysis-ready data,
following a transparent process in which all stages of development
follow open principles \citep{arribas-bel2021}. ODPs, while still open,
differ from general open data in their degree of ease of access, their
heightened usability, and potentially the value they add to the raw
data.

@@

Using Public Use Microdata Files (PUMFs) from the TUS, we extracted all
data necessary to characterize active travel in Canada - specifically,
episodes in which the activity involved moving between an origin and a
destination by walking or cycling. Although Statistics Canada provides
PUMFs and accompanying documentation for the GSS program
\citep[see][]{statisticscanada2024}, accessing and preparing these files
for analysis is not a straightforward task due to their size and
complexity. The process of extracting information of interest from the
source files is time-consuming, tedious, and challenging and/or prone to
error due to the expertise required to work with these files. To create
\{ActiveCA\} we selected, labelled, and processed the TUS cycles to make
them ready for analysis.

@@

Given the level of interest in active travel
\citep[e.g.,][]{mccurdySupport2023}, reducing the barriers to using data
contained in rich, but difficult to access and use surveys, such as TUS,
is a worthy endeavour that can only improve data-driven decisions in
transportation, urban, and health policy.

The rest of this paper discusses the sources of data, and the process
implemented to retrieve and package them. Then, we show some selected
examples of analysis to whet the imagination of potential users. This
ODP provides not only data that are easy to use, but also all the code
and documentation that make this a reproducible research project. In
summary, \{ActiveCA\} aims to implement and inspire the best principles
of open spatial sciences \citep{paez_open_2021, brunsdon_opening_2021}.

@@

Statistics Canada \citeyearpar{statisticscanada2024} conducts GSS
surveys to obtain data on social trends to track changes in Canadians'
living conditions and well-being over time. The series of survey on time
use are used to understand how Canadian residents spend and manage their
time, and what factors contribute to their happiness and stress. The GSS
program was created in 1985, and is serialized to provide a collection
of annual, representative cross-sectional surveys.

The topics of the survey cycle every few years to cover topics that
include family, health, social identity, and every five to seven years
time use. The first Canadian TUS done as part of the GSS program was
conducted in 1986, and the most recent was completed in 2022. These Time
Use Surveys \citep{statisticscanada2022} collect data on respondents'
participation and time spent on a wide range of everyday activities
using a 24-hour retrospective diary, with information on the location of
these activities (e.g.~at home, at work, etc.) and, for non-personal
activities, the people who were present with the respondent at the time
of the activity. In addition, time-use surveys also cover topics related
to leisure time, work-life balance, health, commuting, culture and
sports, and many others.

@@

\begin{table}
\centering
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER: person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

\begin{table}
\centering
\caption{\label{tab:gss-epi-file-2015}\label{tab:ep-2015-unprocessed}Visualization of the  first seven episodes of the record number 10041.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & EPINO & WGHT\_EPI & TUI\_01 & DURATION & LOCATION\\
\midrule
10041 & 1 & 1353.818 & 1 & 210 & 300\\
10041 & 2 & 1353.818 & 2 & 40 & 300\\
10041 & 3 & 1353.818 & 27 & 15 & 300\\
10041 & 4 & 1353.818 & 7 & 15 & 315\\
10041 & 5 & 1353.818 & 9 & 180 & 300\\
\addlinespace
10041 & 6 & 1353.818 & 6 & 15 & 300\\
10041 & 7 & 1353.818 & 18 & 120 & 300\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

Table \ref{tab:processed-obs} displays the total number of records
processed for Main and episode files. For the main files, a total of
101,667 records were processed, referring to all records from the Time
Use Surveys from 1986 to 2022, that together represents more of
181,526,641 respondents. It also presents the total cases of active
trips episodes identified. In total 23,513 records with register of
active travel activity. Together, these records account for 44,316,110
episodes.

\begin{table}
\centering
\caption{\label{tab:table_df_processed}\label{tab:processed-obs}Total number and weighted sum of records processed.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{rr>{}r|rr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Episode} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Year & Count & Weighted & Count\_ep & Weighted\_ep\\
\midrule
2022 & 12336 & 32136802 & 1765 & 6041032\\
2015 & 17390 & 29766399 & 3496 & 6634387\\
2010 & 15390 & 28075610 & 4615 & 8516753\\
2005 & 19597 & 26095819 & 5866 & 7583838\\
1998 & 10749 & 24260137 & 1789 & 3606987\\
\addlinespace
1992 & 9815 & 21294313 & 1635 & 3691918\\
1986 & 16390 & 19897562 & 4347 & 8241196\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the TUS PUMF 2015 Main File (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number \texttt{10041} from the TUS PUMF 2015 episode file
(Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}.
Only the unique active travel episode appears in Table
\ref{tab:ep-2015-processed} since the records were filtered to select
cases with walking or cycling episodes. For both cases, Tables
\ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain
labeled variables, facilitating the interpretation of the data.

\begin{table}
\centering
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10001 & 8516.6140 & July & 55 to 64 years & Male & Married\\
10002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10003 & 1019.3135 & March & 65 to 74 years & Female & Divorced\\
10004 & 1916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10005 & 1952.2015 & April & 15 to 24 years & Male & Single, never married\\
10006 & 5761.5528 & August & 15 to 24 years & Male & Single, never married\\
10007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10008 & 2479.2991 & February & 25 to 34 years & Female & Married\\
10009 & 1436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{tablenotes}
\end{threeparttable}}
\end{table}

\begin{table}
\centering
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number 10041 of the 2015 GSS survey.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{rrlrlll}
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10041 & 1353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

Bruno Dias dos Santos - \url{https://orcid.org/0000-0002-6748-2038}

Mahdis Moghadasi - \url{https://orcid.org/0000-0003-1009-9369}

Antonio Páez - \url{https://orcid.org/0000-0001-6912-9919}

>>>>

\corrauth{Bruno Dias dos Santos - School of Earth, Environment and
Society, McMaster University, 1280 Main St W, Hamilton, Ontario L8S 4K1
Canada.}

@@

The objective of this paper is to introduce \{ActiveCA\}, an open data
product with data from all Time Use Survey (TUS) cycles of the Canadian
General Social Surveys (GSSs). Open data products (ODPs) are the outcome
of a process that transforms raw data (open or not) into analysis-ready
data, following a transparent process in which all stages of development
follow open principles \citep{arribas-bel2021}. ODPs, while still open,
differ from general open data in their degree of ease of access, their
heightened usability, and potentially the value they add to the raw
data.

@@

Using PUMFs from the TUS, we extracted all data necessary to
characterize active travel in Canada - specifically, episodes in which
the activity involved moving between an origin and a destination by
walking or cycling. Although Statistics Canada provides PUMFs and
accompanying documentation for the GSS program
\citep[see][]{statisticscanada2024}, accessing and preparing these files
for analysis is not a straightforward task due to their size and
complexity. The process of extracting information of interest from the
source files is time-consuming, tedious, and challenging and/or prone to
error due to the expertise required to work with these files. To create
\{ActiveCA\} we selected, labelled, and processed the TUS cycles to make
them ready for analysis.

@@

Given the level of interest in active travel
\citep[e.g.,][]{mccurdySupport2023}, reducing the barriers to using data
contained in rich, but difficult to access and use surveys, such as TUS,
is a worthy endeavour that can only improve data-driven decisions in
transportation, urban, and health policy. The rest of this paper
discusses the sources of data, and the process implemented to retrieve
and package them. Then, we show some selected examples of analysis to
whet the imagination of potential users. This ODP provides not only data
that are easy to use, but also all the code and documentation that make
this a reproducible research project. In summary, \{ActiveCA\} aims to
implement and inspire the best principles of open spatial sciences
\citep{paez_open_2021, brunsdon_opening_2021}.

@@

Statistics Canada \citeyearpar{statisticscanada2024} conducts GSS
surveys to obtain data on social trends to track changes in Canadians'
living conditions and well-being over time. TUS are used to understand
how Canadian residents spend and manage their time, and what factors
contribute to their happiness and stress. The GSS program was created in
1985, and is serialized to provide a collection of annual,
representative cross-sectional surveys.

The topics of the survey cycle every few years to cover topics that
include family, health, social identity, and every five to seven years
time use. The first Canadian TUS done as part of the GSS program was
conducted in 1986, and the most recent was completed in 2022. These
surveys \citep{statisticscanada2022} collect data on respondents'
participation and time spent on a wide range of everyday activities
using a 24-hour retrospective diary, with information on the location of
these activities (e.g.~at home, at work, etc.) and, for non-personal
activities, the people who were present with the respondent at the time
of the activity. In addition, time-use surveys also cover topics related
to leisure time, work-life balance, health, commuting, culture and
sports, and many others.

@@

\begin{table}
\centering
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10,000 & 616.6740 & 7 & 5 & 1 & 5\\
10,001 & 8,516.6140 & 7 & 5 & 1 & 1\\
10,002 & 371.7520 & 1 & 4 & 2 & 1\\
10,003 & 1,019.3135 & 3 & 6 & 2 & 5\\
10,004 & 1,916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10,005 & 1,952.2015 & 4 & 1 & 1 & 6\\
10,006 & 5,761.5528 & 8 & 1 & 1 & 6\\
10,007 & 466.0426 & 6 & 5 & 2 & 3\\
10,008 & 2,479.2991 & 2 & 2 & 2 & 1\\
10,009 & 1,436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER: person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

\begin{table}
\centering
\caption{\label{tab:gss-epi-file-2015}\label{tab:ep-2015-unprocessed}Visualization of the  first seven episodes of the record number 10041.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & EPINO & WGHT\_EPI & TUI\_01 & DURATION & LOCATION\\
\midrule
10,041 & 1 & 1,353.818 & 1 & 210 & 300\\
10,041 & 2 & 1,353.818 & 2 & 40 & 300\\
10,041 & 3 & 1,353.818 & 27 & 15 & 300\\
10,041 & 4 & 1,353.818 & 7 & 15 & 315\\
10,041 & 5 & 1,353.818 & 9 & 180 & 300\\
\addlinespace
10,041 & 6 & 1,353.818 & 6 & 15 & 300\\
10,041 & 7 & 1,353.818 & 18 & 120 & 300\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

Table \ref{tab:processed-obs} displays the total number of records
processed for main and episode files. For the main files, a total of
101,667 records were processed, referring to all records from the TUSs
from 1986 to 2022, that together represents more of 181,526,641
respondents. It also presents the total cases of active trips episodes
identified. In total 23,513 records with register of active travel
activity. Together, these records account for 44,316,110 episodes.

\begin{table}
\centering
\caption{\label{tab:table_df_processed}\label{tab:processed-obs}Total number and weighted sum of records processed.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{rr>{}r|rr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Episode} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Year & Count & Weighted & Count\_ep & Weighted\_ep\\
\midrule
2,022 & 12,336 & 32,136,802 & 1,765 & 6,041,032\\
2,015 & 17,390 & 29,766,399 & 3,496 & 6,634,387\\
2,010 & 15,390 & 28,075,610 & 4,615 & 8,516,753\\
2,005 & 19,597 & 26,095,819 & 5,866 & 7,583,838\\
1,998 & 10,749 & 24,260,137 & 1,789 & 3,606,987\\
\addlinespace
1,992 & 9,815 & 21,294,313 & 1,635 & 3,691,918\\
1,986 & 16,390 & 19,897,562 & 4,347 & 8,241,196\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the TUS PUMF 2015 main file (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number \texttt{10041} from the TUS PUMF 2015 episode file
(Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}.
Only the unique active travel episode appears in Table
\ref{tab:ep-2015-processed} since the records were filtered to select
cases with walking or cycling episodes. For both cases, Tables
\ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain
labeled variables, facilitating the interpretation of the data.

\begin{table}
\centering
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{cccccc}
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10,000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10,001 & 8,516.6140 & July & 55 to 64 years & Male & Married\\
10,002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10,003 & 1,019.3135 & March & 65 to 74 years & Female & Divorced\\
10,004 & 1,916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10,005 & 1,952.2015 & April & 15 to 24 years & Male & Single, never married\\
10,006 & 5,761.5528 & August & 15 to 24 years & Male & Single, never married\\
10,007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10,008 & 2,479.2991 & February & 25 to 34 years & Female & Married\\
10,009 & 1,436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{tablenotes}
\end{threeparttable}}
\end{table}

\begin{table}
\centering
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number 10041 of the 2015 GSS survey.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{rrlrlll}
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10,041 & 1,353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

\begin{itemize}
\tightlist
\item
  Bruno Dias dos Santos - \url{https://orcid.org/0000-0002-6748-2038}
\item
  Mahdis Moghadasi - \url{https://orcid.org/0000-0003-1009-9369}
\item
  Antonio Páez - \url{https://orcid.org/0000-0001-6912-9919}
\end{itemize}
